---
title: "Project Planning"
author: "Jenna"
format: html
editor: visual
---

## *NFL Passing Data:*

- Passing data from regular season weeks 1-9 in 2022

- Lacks target player position & all players' headshots

## *Question Breakdown:*

1.  What's the most successful play on fourth down?

    *Variables - Total Yards, Completion Percentage, Route Ran, Huddle/No Huddle, Offense Formation, Motion Since Lineset, Time to Throw*

2.  Who were the top passing/receiving duos?

    *How often were these receivers targeted by their QB?*

```{r}
### Exploring Data ----------------------------------------------------------------

# Loading libraries
library(tidyverse)
library(nflreadr)
library(ggimage)
library(teamcolors)
library(sportyR)
library(imager)
library(patchwork)
library(dplyr)

# Inspect data
nfl_passes <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/nfl_passes.csv")

head(nfl_passes)
summary(nfl_passes)
```

```{r}
### Data Tidying ------------------------------------------------------------------

# Source player headshots and position from nflreadr package
player_headshots <- nflreadr::load_rosters(seasons = 2022) 
player_headshots <- player_headshots|>
  select(position, full_name, headshot_url)

# Join position and target player headshots to dataset
df <- nfl_passes |>
  left_join(player_headshots, by = join_by("target_player_name" == "full_name"), relationship = "many-to-many")

# Generate dataframe to prepare joining headshots of quarterbacks
quarterbacks <- player_headshots |>
  mutate(full_name = str_replace(full_name, "^(\\w)\\w*\\s+", "\\1.")) |>
  select(full_name, qb_headshot = headshot_url)
```

```{r}
nfl_passes |>
  filter(down == 4, !is.na(route_ran)) |>
  ggplot(aes(x = target_x, target_y, color = complete_pass)) +
  geom_point() +
  facet_wrap(~route_ran) +
  coord_flip() +
  theme()
```

```{r}
# Generating dataset with completed passes during fourth down routes
fourth_down_outcomes <- df |>
  filter(!is.na(position),
         position %in% c("RB", "TE", "WR"),
         down == "4"
         ) |>
  group_by(route_ran, position) |>
  summarise(avg_air_yards = mean(air_yards), avg_total_yards = mean(yards_gained), completed_pass_perc = mean(complete_pass))

# Dataset sample
head(fourth_down_outcomes)
```

```{r}
# Generates graph to illustrate total yards during fourth down completions based on different routes and positions
fourth_down_outcomes |>
  mutate(expected_yards = avg_total_yards * completed_pass_perc) |>
  ggplot(aes(x = route_ran, y = expected_yards)) +
  geom_col() +
   facet_wrap(~position) +
  coord_flip() 


fourth_down_outcomes |>
  mutate(expected_yards = avg_total_yards * completed_pass_perc) |>
  ggplot(aes(y = route_ran, x = position)) +
  geom_tile(aes(fill = expected_yards), color = "grey50") +
  scale_fill_continuous() +
  theme_classic() +
  coord_equal()
```

```{r}
#
summary(nfl_passes$target_x)
summary(nfl_passes$target_y)

# Constructing football field
football_field <- geom_football(league = "NFL", display_range = "in_bounds_only", rotation = 270)

# Fourth down plays
fourth_downs <- nfl_passes |>
  filter(down == 4, !is.na(route_ran)) |>
  group_by(route_ran) |>
  mutate(attempt = n()) |>
  filter(attempt > 10) |>
  ungroup() 

# 
football_field + 
  geom_point(data = fourth_downs, 
             aes(x = target_y -26.65, y = target_x - 60, color = factor(complete_pass))) + 
  facet_wrap(~route_ran) +
  scale_color_manual(
    values = c("0" = "#D9381E", "1" = "lightgreen"), 
    labels = c("0" = "Incomplete", "1" = "Complete"),
    name = "Pass Outcome"
  ) +
  theme_classic()

```

```{r}
# Clustering
fourth_down_features <- df |>
  mutate(no_huddle = as.numeric(no_huddle),
         complete_pass = as.numeric(complete_pass),
         qtr = as.numeric(qtr),
         offense_formation = as.factor(offense_formation)) |>
  select(yards_gained, 
         no_huddle, 
         qtr, 
         time_to_throw,
         complete_pass, 
         ) |> 
  drop_na() 

std_fourth_down_features <- fourth_down_features |> 
  scale(center = TRUE, scale = TRUE)

set.seed(47)
kmeans_many_features <- fourth_down_features |> 
  kmeans(algorithm = "Hartigan-Wong", centers = 4, nstart = 30) 
```

```{r}
# Clustering time_to_throw and yards_gained, standardized
set.seed(47)

clean_football_features <- fourth_downs |>
  filter(complete_pass == 1) |>
  mutate(std_time_to_throw = as.numeric(scale(time_to_throw, center = TRUE, scale = TRUE)), 
         std_yards_gained = as.numeric(scale(yards_gained, center = TRUE, scale = TRUE)))

std_kmeans <- clean_football_features |> 
  select(std_time_to_throw, std_yards_gained) |> 
  kmeans(algorithm = "Lloyd", centers = 4, nstart = 30, iter.max = 50)

clean_football_features |>
  mutate(
    team_clusters = as.factor(std_kmeans$cluster)
  ) |>
  ggplot(aes(x = std_time_to_throw, y = std_yards_gained,
             color = team_clusters)) +
  geom_point(size = 2) + 
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom") +
  coord_fixed() +
  facet_wrap(~route_ran)


fourth_downs |>
  filter(complete_pass ==1) |>
  summarise(mean(yards_gained))
```

```{r}
# Clustering
init_kmeans <- fourth_downs |>
  filter(complete_pass == 1) |>
  select(time_to_throw, yards_gained) |>
  kmeans(algorithm = "Lloyd", centers = 4,
         nstart = 30, iter.max = 50)

# Not standardized, causes kills per set to impact more than blocks per set
fourth_downs |>
  filter(complete_pass == 1) |>
  mutate(team_clusters = as.factor(init_kmeans$cluster)) |>
  ggplot(aes(x = time_to_throw, y = yards_gained,
             color = team_clusters)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "bottom") +
  coord_fixed() + 
  facet_wrap(~motion_since_lineset)
```

```{r}

test <- df |>
  group_by(passer_player_name, target_player_name, receiver_headshot = headshot_url, posteam) |>
  summarise(times_targeted = n(), 
            total_yards = sum(yards_gained), 
            yards_per_target = total_yards/times_targeted, 
            .groups = "drop") |>
  arrange(desc(total_yards)) |>
  filter(times_targeted >= 5) |>
  head(10)

test |>
  ggplot(aes(x = times_targeted, y = total_yards)) +
  geom_image(aes(image = receiver_headshot), size = 0.2) 

test2 <- test |>
  ggplot(aes(x = reorder(target_player_name, total_yards), y = total_yards, fill = posteam)) +
  geom_col(width = 0.52) +
  geom_image(aes(image = receiver_headshot), size = 0.16) +
  coord_flip() +
  theme_classic()

top_duos <- df |>
  left_join(quarterbacks, by = join_by("passer_player_name" == "full_name"), relationship = "many-to-many") |>
  group_by(passer_player_name, 
           target_player_name, 
           receiver_headshot = headshot_url, 
           qb_headshot,
           posteam) |>
  summarise(times_targeted = n(), 
            total_yards = sum(yards_gained), 
            yards_per_target = total_yards/times_targeted,
            .groups = "drop") |>
  distinct(passer_player_name, target_player_name, .keep_all = TRUE) |>
  arrange(desc(total_yards))|>
  head(10)

# Team colors from top 10 duos
teamcolors <- c(
  "LA" = "#0C2BE9",
  "MIN" = "#4f2683",
  "MIA" = "#10b8bd",
  "BUF" = "#0d48a2",
  "KC" = "#e31837",
  "PHI" = "#004C54",
  "TB" = "#c60c30",
  "JAX" = "#006778",
  "CIN" = "#fb4f14",
  "CLE" = "#7c4700",
  "LV" = "grey")

duo_attempts <- df |>
  group_by(passer_player_name) |>
  mutate(qb_attempts = n()) |>
  ungroup() |>
  group_by(passer_player_name, target_player_name) |>
  mutate(count = n()) |>
  distinct(passer_player_name, target_player_name, count, qb_attempts)

top_duos |>
  left_join(duo_attempts, by = c("passer_player_name", "target_player_name")) |>
  mutate(percent_targeted = 100 * round(count/qb_attempts, 2),
         target_player_name = str_replace(target_player_name, "^(\\w)\\w*\\s+", "\\1."), duo = paste(passer_player_name, target_player_name, sep = " & "), 
         metrics = paste0(total_yards, " (", percent_targeted, "%)")) |>
  ggplot(aes(x = reorder(duo, total_yards), 
             y = total_yards, 
             fill = posteam)) +
  geom_col(width = 0.6) +
  geom_image(aes(image = receiver_headshot), 
             size = 0.16) +
  geom_image(aes(y = 0, image = qb_headshot), 
             size = 0.16) +
  geom_text(aes(label = duo),
            family = "mono",
            fontface = "bold", 
            y = 35, 
            color = "white", 
            hjust = 0) +
  geom_text(aes(label = metrics),
            family = "mono",
            fontface = "bold",
            color = "white",
            hjust = 1.3) +
  scale_y_continuous(breaks = seq(0, 1000, by = 200)) +
  coord_flip() +
  scale_fill_manual(values = teamcolors) +
  labs(title = "Top Passing Duos in the NFL",
       subtitle = "From the 2022 Season",
       y = "Total Yards (Perc. Targeted)",
       x = "") +
  theme_classic() +
  theme(legend.position = "none",
        plot.title = element_text(size = 25, family = "mono"),
        plot.subtitle = element_text(size = 14, family = "mono"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

```
